---
title: 'Maps with Leaflet'
author: "Emmanuel Ikekwere"
date: "2023-04-30"
output: 
  html_document:
    toc: true
    toc_float: TRUE
    code_folding: hide
---

# Introduction

### Landfill Methane Outreach Program (LMOP)
The LMOP Landfill and Landfill Gas Energy Project Database (LMOP Database) contains information such as physical address, latitude and longitude, owner/operator organization, operational status, year opened, actual or expected closure year, design capacity, amount of waste in place, gas collection system status and landfill gas (LFG) collected amount for 2,637 municipal solid waste landfills. For landfills that report under EPA’s Greenhouse Gas Reporting Program (GHGRP), LMOP cross-references that data set by including GHGRP’s 7-digit Facility Identifier.

Aim: The main goal of this research is to utilize the map to analyze locations and characteristics of landfills in the United States. 

Data: The data are available from the Environmental Protection Agency (EPA) as referenced here. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)

library(dplyr)
library(ggplot2)
library(readxl)
library(expss)
library(sp)
library(rgeos)
library(rgdal)
library(leaflet)
library(mapview)

setwd("C:/Users/ADMIN/Desktop/ANLY/ANLY_512_DV/Maps with Leaflet Assignment")
```
# Dataset

"The LMOP Database contains key information about MSW landfills and LFG energy projects in the United States. Information in the LMOP Database is compiled from a variety of sources. Data received for inclusion in the LMOP Database are reviewed for reasonableness and are corroborated via other data sources when possible. Not all data are updated each year. While the Program strives to keep the information as updated and accurate as possible, the database is not exhaustive. The LMOP Database does not include data for every MSW landfill in the United States." - EPA

```{r pressure, echo=FALSE}
# Reading the excel file sheet where the dataset is contained
landfilldata <- read_excel("landfilllmopdata.xlsx", sheet = "LMOP Database")

# Selecting and renaming variables
landfilldata <- landfilldata %>%
  select("Landfill Name",
         City,
         County,
         State,
         Latitude,
         Longitude,
         "Current Landfill Status",
         "Landfill Design Capacity (tons)",
         "Waste in Place (tons)",
         "Annual Waste Acceptance Year",
         "Annual Waste Acceptance Rate (tons per year)") %>% 
  rename(name = "Landfill Name",
         city = City,
         county = County,
         state = State,
         latitude = Latitude,
         longitude = Longitude,
         c_status = "Current Landfill Status",
         land_design_cap = "Landfill Design Capacity (tons)",
         waste_in_place = "Waste in Place (tons)",
         waste_accp_year = "Annual Waste Acceptance Year",
         waste_annual_accp = "Annual Waste Acceptance Rate (tons per year)")

# factoring the city, county, state and current status. 
landfilldata$city <- factor(landfilldata$city)
landfilldata$county <- factor(landfilldata$county)
landfilldata$state <- factor(landfilldata$state)
landfilldata$c_status <- factor(landfilldata$c_status)
str(landfilldata)
summary(landfilldata)

```

# Data Summary

Using the 'expss' function to generate baseline statistical tabulations of the data. We generate tabulations to display the state as well as three (3) other waste indicators based on the current status (open, closed). The map can then be used to analyze the data.

```{r}
landfilldata %>% 
  tab_cols(total(), c_status) %>%
  tab_cells(state) %>%
  tab_stat_cases() %>%
  tab_pivot() %>%
  set_caption("State Statistics in Summary") %>%
  htmlTable(., css.cell = c("width: 275px", rep("width: 95px", ncol(.) - 1)))


landfilldata %>% 
  tab_cols(total(), c_status) %>%
  tab_cells(land_design_cap, waste_in_place, waste_annual_accp) %>%
  tab_stat_mean_sd_n() %>%
  tab_pivot() %>%
  set_caption("Wastes Statistics in Summary")%>%
  htmlTable(., css.cell = c("width: 275px", rep("width: 84px", ncol(.) - 1)))
```

# Annual Waste Acceptance Rate (tons) for United States in 2021. 

Looking at each state's Annual Waste Acceptance Rate, which is a critical indicator of landfill treatment capacity based on EPA data in 2021. Based on this data, the states with the highest pressure for trash recycling are California, Texas, regions in the southeast, northeast, and the Great Lakes.

# Maps with Leaflet
```{r}

setwd("C:/Users/ADMIN/Desktop/gisData")
states <- readOGR(dsn=".", layer="cb_2021_us_state_500k", verbose = FALSE )

```
```{r}
states_center <- data.frame(gCentroid(states, byid = TRUE))
states$lng <- states_center$x
states$lat <- states_center$y

landfilldata_state_accp_sum <- landfilldata %>% filter(waste_accp_year == 2021) %>% group_by(state) %>% summarise(accp_sum = sum(waste_annual_accp, na.rm = TRUE))
sp_landfilldata_state_accp_sum <- merge(states, landfilldata_state_accp_sum, by.x="STUSPS", by.y="state")

```

## Map 1

This leaflet map provides an overview of the waste acceptance rate in 2021 across the United States. In order to present the data, it has been divided into color bins with values ranging from 0 to 35,000,000 tons. Selecting various states will allow the user to view the specific acceptance rates for each state. Additionally, there is also a legend provided to indicate the color-coded values that represent different waste acceptance rates.

```{r}
bins <- c(0, 100000, 1500000, 3500000, 7500000, 35000000, Inf)
pal <- colorBin("RdYlGn", domain = sp_landfilldata_state_accp_sum$accp_sum, bins=bins)
sp_landfilldata_state_accp_sum$label <- paste(sp_landfilldata_state_accp_sum$NAME,":", sp_landfilldata_state_accp_sum$accp_sum, "(tons)")

#Let's add the leaflet map
map_1 <- leaflet(sp_landfilldata_state_accp_sum) %>%
  addTiles() %>%
  addProviderTiles(provider="OpenStreetMap") %>%
  setView(-98, 38.5, 4) %>%
  addPolygons(fillColor = ~pal(accp_sum),
              fillOpacity = 0.7,
              weight = 1.5,
              dashArray = "3",
              color = "#ffffff",
              opacity = 2,
              highlightOptions = highlightOptions(weight = 2.5,
                                           color = "#AAA",
                                           dashArray = "",
                                           fillOpacity = 0.7,
                                           bringToFront = TRUE),
              label = ~label,
              labelOptions = labelOptions(style = list(padding = "6px"), textsize = "13px")
  )%>% 
  addMarkers(lng = ~lng, 
             lat = ~lat,
             label = ~STUSPS,
             labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE)
  )%>%
  addLegend(pal = pal, opacity = 0.7, values = ~accp_sum, position = "bottomright", title = "Waste Acceptance Rate 2021")


map_1
```
# US States Landfill Remaining Capacity (tons). 
In addition to the Annual Waste Acceptance Rate, Landfill Remaining Capacity is another key indicator for displaying reserve capacity. As you can see, each state has a different amount of landfill capacity available for disposal. Based on the following map, it can be seen that Texas, California, and some southwestern states have relatively strong remaining capacity.

## Map 2
```{r warning=FALSE, message=FALSE}
# US States Landfill Remaining Capacity (tons)
landfilldata_state_re_sum <- landfilldata %>% filter(c_status == "Open") %>%
  group_by(state) %>% summarise(re_sum = sum(land_design_cap - waste_in_place, na.rm = T))
sp_landfilldata_state_re_sum <- merge(states, landfilldata_state_re_sum, by.x="STUSPS", by.y="state")

bins <- c(0, 30000000, 150000000, 450000000, 1500000000, Inf)
mypal <- colorBin("RdBu", domain = sp_landfilldata_state_re_sum$re_sum, bins = bins)
sp_landfilldata_state_re_sum$label <- paste(sp_landfilldata_state_re_sum$NAME, ":", sp_landfilldata_state_re_sum$re_sum, "(tons)")

#Let's add the leaflet map

map_2 <- leaflet(sp_landfilldata_state_re_sum) %>%
  addTiles() %>%
  setView(-98, 38.5, 4) %>%
  addProviderTiles(provider="CartoDB") %>%
  addPolygons(fillColor = ~mypal(re_sum),
              fillOpacity = 0.7,
              weight = 1.5,
              color = "#ffffff",
              opacity = 1,
              highlightOptions = highlightOptions(weight = 3,
                                                  color = "#AAA",
                                                  fillOpacity = 0.7,
                                                  bringToFront = TRUE),
              label = ~label,
              labelOptions = labelOptions(style = list(padding = "6px"), textsize = "13px")
  )%>%
  addLabelOnlyMarkers(lng = ~lng, 
                      lat = ~lat, 
                      label = ~STUSPS,
                      labelOptions = labelOptions(noHide = TRUE, direction = 'center', textOnly = TRUE)
                      
  )%>%
  addLegend(pal = mypal, opacity = 0.7, values = ~re_sum, position = "bottomright", title = "Landfill Remaining Capcity")


map_2
```
# Landfill sustainability (years) in the United States
Although each state has a different population and other factors, we need another indicator to represent landfill capacity more accurately. Using the landfill sustainability, we are able to determine how long a state would be able to satisfy its landfill requirements without having to build new sites. 

## Map 3
```{r}
landfilldata_state_accp_re <- merge(landfilldata_state_accp_sum, landfilldata_state_re_sum, by.x = "state", by.y= "state")

landfilldata_state_accp_re$sustainable <- round(landfilldata_state_accp_re$re_sum /landfilldata_state_accp_re$accp_sum)
sp_landfilldata_state_accp_re <- merge(states, landfilldata_state_accp_re, by.x= "STUSPS", by.y="state")

sp_landfilldata_state_accp_re$label <- paste(sp_landfilldata_state_accp_re$NAME, ":",sp_landfilldata_state_accp_re$sustainable, "(years)")

#Let's add the leaflet map
map_3 <- leaflet(sp_landfilldata_state_accp_re) %>%
  addTiles() %>%
  setView(-98, 38.5, 4) %>%
  addProviderTiles(provider="CartoDB") %>%
  addPolygons(fillColor = "#006837",
              fillOpacity = 0.7,
              weight = 1.5,
              color = "#ffffff",
              opacity = 1.0) %>%
  addCircles(data = sp_landfilldata_state_accp_re,
             lng = ~lng,
             lat = ~lat,
             radius = ~sustainable*900,
             weight = 1.5, 
             color = "#FFFFE5",
             opacity = 0.8,
             fillColor = "#FFFFE5",
             fillOpacity = 0.8,
             label = ~label,
             highlightOptions = highlightOptions(weight = 3,
                                                 color = "#AAA",
                                                 opacity = 1.0,
                                                 bringToFront = TRUE),
             labelOptions = labelOptions(style = list(padding = "6px"), textsize = "13px")
  ) 


map_3
```
## Summary
As a result of the assessment, this describes the two maps which were distinctively generated to indicate each state's sustainability. The first indicates that Nevada and Utah have great sustainability, whereas the second shows that New England region, particularly Massachusetts, Rhode Island, and Vermont are experiencing severe problems with refuse disposal.


## Map 4
To answer the question about using a rectangle layer to emphasize an area. This map below illustrates a rectangle layer to emphasize the  with severe problems, refuse disposal and less sustainability with the data provided by the Environmental Protection Agency (EPA).  
```{r}
# Map 4
map_4 <- leaflet(sp_landfilldata_state_accp_re) %>%
  addTiles() %>%
  setView(-84.6, 40.7, 4) %>%
  addProviderTiles(provider="OpenStreetMap") %>%
  addPolygons(fillColor = "#006837",
              fillOpacity = 0.6,
              weight = 1.5,
              color = "#ffffff",
              opacity = 0.7) %>%
  addCircles(data = sp_landfilldata_state_accp_re,
             lng = ~lng,
             lat = ~lat,
             weight = 3,
             color = "Orange",
             opacity = 0.6,
             fillColor = "Orange",
             fillOpacity = 0.6,
             label = ~label,
             highlightOptions = highlightOptions(weight = 6,
                                                 color = "#AAA",
                                                 opacity = 1,
                                                 bringToFront = TRUE),
             labelOptions = labelOptions(style = list(padding = "6px"), textsize = "14px")
  )%>%
  addRectangles(-73.03, 44.939108, -70.45, 41.347020,
                fillColor = "transparent") %>% # adding rectangles to states coordinate
  # Vermont, Massachusetts, Rhode Island, and  Connecticut.
  addMiniMap()

map_4

```

# Texas Landfill sustainability in (years)

We will then narrow the scope of our analysis to the situation in Texas, where I live, and analyze it below. In this case, landfill environmental sustainability will be used as an example. Each of the counties shown here decribes the size of their sustainability up until 2021. Few counties has the highest waste disposal facilities sustainability in Texas (TX), along with more than quarter not listed among the counties in TX having adequate landfill sustainability. 

```{r}
setwd("C:/Users/ADMIN/Desktop/gisData")
counties <- readOGR(dsn=".", layer="cb_2016_us_county_500k", verbose = FALSE)
#counties (pre-processed from cb_2016_us_county)
TX_counties <- counties[counties$STATEFP==48,]  # 48 is the FIPS code for Texas

# Now,let's calculate the centroid of the counties in TX and then assign the longitude and latitude values of the county. 
TX_counties_center <- data.frame(gCentroid(TX_counties, byid = TRUE))
TX_counties$lng <- TX_counties_center$x
TX_counties$lat <- TX_counties_center$y

# To filter the landfill data set to only include landfills that are located in Texas and have open status. Then, we can calculate the sustainability of each landfill by subtracting the amount of waste in place from the land design capacity and divide it by annual acceptance rate. 
landfilldata_TX_open <- landfilldata %>% filter(state=="TX", c_status=="Open")
landfilldata_TX_open$sustainable <- round((landfilldata_TX_open$land_design_cap - landfilldata_TX_open$waste_in_place) / landfilldata_TX_open$waste_annual_accp)
landfilldata_TX_open$label <- paste(landfilldata_TX_open$name, ": ", landfilldata_TX_open$sustainable, "(years)")

# Adding more commands to make the map

leaflet(TX_counties) %>%
  addTiles() %>%
  setView(-95.4419, 29.55, 7) %>%
  addPolygons(weight = 1.5, 
              opacity = 1,
              fillColor = "#e4b849", 
              color = "#ffffff", 
              fillOpacity = 0.7) %>%
  addCircles(data = landfilldata_TX_open,
             lng = ~longitude,
             lat = ~latitude,
             radius = ~log(sustainable, 2) * 1200,
             weight = 2,
             opacity = 0.6,
             color = "red",
             fillColor = "#00ff00",
             fillOpacity = 0.6,
             label = ~label,
             highlight = highlightOptions(weight = 6,
                                          color = "#AAA",
                                          opacity = 1,
                                          bringToFront = TRUE),
             labelOptions = labelOptions(style = list(padding = "6px"), textsize = "13px")
             )%>%
  addLabelOnlyMarkers(lng = ~lng,
                      lat = ~lat,
                      label = ~NAME,
                      labelOptions = labelOptions(style = list("color"="black", "font-size" = "13px"), noHide = TRUE, direction = 'center', textOnly = TRUE)
                      )

```


## Narrowing coordinates to the county of my residence - Williamson

```{r}
leaflet(TX_counties) %>%
  addTiles() %>%
  setView(-97.55400, 30.60200, 9) %>%
  addPolygons(weight = 1.5, 
              opacity = 1,
              fillColor = "#e4b849", 
              color = "#ffffff", 
              fillOpacity = 0.7) %>%
  addCircles(data = landfilldata_TX_open,
             lng = ~longitude,
             lat = ~latitude,
             radius = ~10000/log(sustainable + 2), #Calculates the radius of the circles in the leaflet map based on the sustainability of the landfill data.
             weight = 2,
             opacity = 0.6,
             color = "red",
             fillColor = "#00ff00",
             fillOpacity = 0.6,
             label = ~label,
             highlight = highlightOptions(weight = 6,
                                          color = "#AAA",
                                          opacity = 1,
                                          bringToFront = TRUE),
             labelOptions = labelOptions(style = list(padding = "6px"), textsize = "13px")
             )%>%
  addLabelOnlyMarkers(lng = ~lng,
                      lat = ~lat,
                      label = ~NAME,
                      labelOptions = labelOptions(style = list("color"="black", "font-size" = "13px"), noHide = TRUE, direction = 'center', textOnly = TRUE)
                      )

```

#Landfill within Round Rock 


```{r}
library(geosphere)

landfilldata_TX_RR_open <- landfilldata %>% filter(state=="TX", c_status=="Open")
landfilldata_TX_RR_open$label <- paste(landfilldata_TX_RR_open$name)
src_lng <- -97.55400
src_lat <- 30.60200

src_gis <- rbind(c(src_lng, src_lat))
dest_gis <- cbind(landfilldata_TX_RR_open$longitude, landfilldata_TX_RR_open$latitude)

dist<- cbind(c(distm(src_gis, dest_gis)))

line <- landfilldata_TX_RR_open %>% filter(dist == min(dist)) %>% 
  select(longitude, latitude)
line <- add_row(line, longitude = src_gis[1], latitude = src_gis[2])

landfilldata_TX_RR_open$popup_content <- paste("<center style = 'font-size: 13px; color:green;'>", landfilldata_TX_RR_open$name, "</center>")

williamson <- leaflet(landfilldata_TX_RR_open) %>%
  addTiles() %>%
  setView(src_lng, src_lat, 8) %>%
  addMarkers(lng = ~longitude,
             lat = ~latitude,
             popup = ~popup_content,
             options = popupOptions(closeButton = FALSE),
             #options = popupOptions("C:\Users\ADMIN\Desktop\NeighbourhoodTrash.jpg")
  ) %>%
  
  addPopups(src_lng, src_lat, ~popup_content, options = popupOptions(closeButton = FALSE))

williamson



```
# Results
We can deduce that the states of Texas, California, Northeast, Southeast, and the Great Lakes have the most trash. 

In comparison to other states, Utah and Nevada are thought to have the best landfill sustainability.

Texas and other southwestern states have a relatively large landfill capacity.

Vermont, Rhode Island, Massachusetts, and the New England region have been identified to experience severe landfill issues and a lack of refuse disposal.  

Blue Ridge LF in Fort Bend County, which has the highest landfill design capacity in Texas, demonstrates a strong commitment to sustainability when the large amount of waste in place (tons) is also factored.  

The Austin Community RDF in Travis County is the closest Waste Management facility to my home.



# References

Cohen, B. (2023). Lecture on Leaflet - Harrisburg University of Science and Technology


Leaflet for R. (n.d.). Retrieved May 1, 2023, from https://rstudio.github.io/leaflet/basemaps.html


The data are available from the Environmental Protection Agency (EPA). Retrieved April 26, 2023. https://www.epa.gov/lmop/landfill-technical-data. link named "Landfill-level data only.


Mercer County. (n.d.). State FIPS Code Listing. Retrieved April 26, 2023, from https://www.mercercountypa.gov/dps/state_fips_code_listing.htm


U.S. Census Bureau. (2023). US Shape Files. Retrieved April 29, 2023, from https://www.census.gov/geographies/mapping-files/2021/geo/carto-boundary-file.html


Legend and color of polygons in R from https://stackoverflow.com/questions/56659000/the-legend-colors-are-not-perfectly-similar-to-color-of-polygons